### 2010-10-22 Thesis/Chapter5.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{The total Lagrangian explicit dynamics algorithm}
\label{chap5}
\begin{shortAbstract}
In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. Their work constituted a step towards the simulation of entire organs in real-time. Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method. However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation. \cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind. We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well. This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}. This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions. Their GPU implementations will be the object of the next chapter. 
\end{shortAbstract}

@@

The first key idea of the TLED algorithm is to refer all variables to the underformed configuration. This type of formulation is said to be \emph{total Lagrangian}. Because of this, the choice of the second Piola-Kirchhoff stress tensor as a measure of stress is required. The strain measure that is work-conjugate with the second Piola-Kirchhoff stress is the Green-St.Venant strain tensor. That is,  the work of stress increments on the strain increments gives an accurate expression for work \citep{Ji10}. We should note that all derivatives in the definition of the Green tensor are with respect to the initial and undeformed configuration. The decisive advantage is that all derivatives are calculated with respect to the undeformed configuration and therefore can be precomputed. This is at the cost to a more complex strain-displacement matrix due an initial displacement effect in the incremental strain \citep{Bathe95}. However, the TLED algorithm performs significantly fewer mathematical operations in each time step.

@@

		\subsubsection*{Constitutive equations and evaluation of stress}
We know that the stresses result from the deformation of the material and they may be expressed in terms of some measure of this deformation such as the strain. The constitutive equations, which depends on the material under consideration, relate the stresses to the strain. As we have seen in section \ref{chap2:elasticity}, hyperelastic materials are a general class of materials in which the constitutive relationship is expressed in the form of a strain energy density function $ _0^t W $. In such materials, the stresses are obtained by differentiating this energy density function with respect to the appropriate strain measure (that is energically conjugate). In a total Lagrangian framework, the appropriate stress and strain measures are second Piola-Kirchhoff stress $ _0^t \mathbf{S} $ and Green-Lagrange strain $ _0^t \mathbf{E} $. Consequently, the stress may be computed from:
\begin{equation}
_0^t\mathbf{S} = \pd{ \leftidx{_0^t}{ \mathbf{W} }{} }{_0^t \mathbf{E}}.
\end{equation}

@@

Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints\label{chap5:discussionTimestep}. As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $. Essentially, stability requires that the time step be not larger than the time required for a dilatational wave to propagate through the smallest element. In linear analyses this translates into the following formula for $ \Delta t_{cr}$:
\begin{equation}
\label{chap5:criticalTimestep}
\Delta t_{cr} = \dfrac{L_e}{c},
\end{equation}
where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material, given by
\begin{equation}
\label{chap5:waveSpeed}
c = \sqrt{\dfrac{E (1 - \nu)}{\rho (1+\nu)(1-2\nu)}},
\end{equation}
where $E$ is Young's modulus and $\nu$ is Poisson's ratio. If nonlinear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase. Additionally, the presence of damping necessitates a further decrease in the limit. Nonetheless, \eqref{chap5:waveSpeed} may be used to estimate $ \Delta t_{cr}$.

@@

	\subsection{Recapitulation}	
For a visco-hyperelastic material stress may be obtained from \eqref{chap5:stress}. This form is general in the sense that any underlying hyperelastic response may be used, including the anisotropic formulations described \eqref{chap5:IstrainEnergy}, supplemented with \eqref{chap5:Phi_vol}, \eqref{chap5:Phi_iso}, and \eqref{chap5:PhiBar} is the general form of an isotropic hyperelastic stress response, defined in terms of invariants. Transversely isotropic or orthotropic models may be produced by substituting \eqref{chap5:PhiBar} for \eqref{chap5:PhiBarTI} or \eqref{chap5:PhiBarOrtho}, respectively. The specification of particular forms of  $ \Psi^{vol} $ and $ \Psi^{iso} $ would be motivated by the particular tissue/material under analysis, and may stem from phenomenological or microstructural considerations. Finally we note that for separated isochoric and volumetric hyperelastic functions as used here, viscoelastic terms may be applied to either or both independently.

@@

>>>>

\chapter{The total Lagrangian explicit dynamics algorithm}
\label{chap5}
\begin{shortAbstract}
In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well. A crucial advantage of this enhanced TLED algorithm is that the computations are conducted at the element level and therefore the algorithm can be easily parallelised. This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation. This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions. Their GPU implementations will be the object of the next chapter. 
\end{shortAbstract}

@@

The first key idea of the TLED algorithm is to refer all variables to the undeformed configuration. This type of formulation is said to be \emph{total Lagrangian}. Because of this, the choice of the second Piola-Kirchhoff stress tensor as a measure of stress is required. The strain measure that is work-conjugate with the second Piola-Kirchhoff stress is the Green-St.Venant strain tensor. That is,  the work of stress increments on the strain increments gives an accurate expression for work \citep{Ji10}. We should note that all derivatives in the definition of the Green tensor are with respect to the initial and undeformed configuration. The decisive advantage is that all derivatives are calculated with respect to the undeformed configuration and therefore can be precomputed. This is at the cost to a more complex strain-displacement matrix due an initial displacement effect in the incremental strain \citep{Bathe95}. However, the TLED algorithm performs significantly fewer mathematical operations in each time step.

@@

		\subsubsection*{Constitutive equations and evaluation of stress}
We know that the stresses result from the deformation of the material and they may be expressed in terms of some measure of this deformation such as the strain. The constitutive equations, which depends on the material under consideration, relate the stresses to the strain. As we have seen in section \ref{chap2:elasticity}, hyperelastic materials are a general class of materials in which the constitutive relationship is expressed in the form of a strain energy density function $ _0^t W $. In such materials, the stresses are obtained by differentiating this energy density function with respect to the appropriate strain measure (that is energetically conjugate). In a total Lagrangian framework, the appropriate stress and strain measures are second Piola-Kirchhoff stress $ _0^t \mathbf{S} $ and Green-Lagrange strain $ _0^t \mathbf{E} $. Consequently, the stress may be computed from:
\begin{equation}
_0^t\mathbf{S} = \pd{ \leftidx{_0^t}{ \mathbf{W} }{} }{_0^t \mathbf{E}}.
\end{equation}

@@

Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints\label{chap5:discussionTimestep}. As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $. Essentially, stability requires that the time step be not larger than the time required for a dilatational wave to propagate through the smallest element. In linear analyses this translates into the following formula for $ \Delta t_{cr}$:
\begin{equation}
\label{chap5:criticalTimestep}
\Delta t_{cr} = \dfrac{L_e}{c},
\end{equation}
where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material, given by
\begin{equation}
\label{chap5:waveSpeed}
c = \sqrt{\dfrac{E (1 - \nu)}{\rho (1+\nu)(1-2\nu)}},
\end{equation}
where $E$ is Young's modulus and $\nu$ is Poisson's ratio. If non-linear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase. Additionally, the presence of damping necessitates a further decrease in the limit. Nonetheless, \eqref{chap5:waveSpeed} may be used to estimate $ \Delta t_{cr}$.

@@

	\subsection{Recapitulation}	
For a visco-hyperelastic material stress may be obtained from \eqref{chap5:stress}. This form is general in the sense that any underlying hyperelastic response may be used, including the anisotropic formulations described \eqref{chap5:IstrainEnergy}, supplemented with \eqref{chap5:Phi_vol}, \eqref{chap5:Phi_iso}, and \eqref{chap5:PhiBar} is the general form of an isotropic hyperelastic stress response, defined in terms of invariants. Transversely isotropic or orthotropic models may be produced by substituting \eqref{chap5:PhiBar} for \eqref{chap5:PhiBarTI} or \eqref{chap5:PhiBarOrtho}, respectively. The specification of particular forms of  $ \Psi^{vol} $ and $ \Psi^{iso} $ would be motivated by the particular tissue/material under analysis, and may stem from phenomenological or micro-structural considerations. Finally we note that for separated isochoric and volumetric hyperelastic functions as used here, viscoelastic terms may be applied to either or both independently.

@@

\section{Conclusion}

The main advantage of the TLED is that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation \citep{Miller07}. Coupled with the use of an explicit integration scheme that eliminates the need for an iterative equation solving, the work of \cite{Miller07} constituted a step towards the simulation of entire organs in real-time. However, explicit analyses impose a critical time step for the simulation to remain stable (see detailed discussion about the critical time step in section~\ref{chap6:criticalTimestep} page~\pageref{chap6:criticalTimestep}).

A constitutive update procedure for anisotropic and visco-hyperelastic materials has also been presented. The procedure allows the incorporation of these elaborate mechanical properties with only a small additional computational cost.

A crucial benefit of this enhanced TLED algorithm is that the computations are conducted at the element level and therefore the algorithm can be easily parallelised. Hence, in 2007 \citeauthor{Taylor07b} re-formulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method. However, restrictions due to using a graphics-based API necessitated re-formulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind. We also improved the model by including the anisotropic and viscoelastic features in the GPU implementation. This implementation is detailed in the following chapter.

### 2010-10-22 Thesis/Chapter6.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation. If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation. The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation. This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA. This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
\end{shortAbstract}

\section{Summary of the TLED formulation}
A complete description of the TLED algorithm for soft tissue simulation was given in the previous chapter. However, let us remind the main steps of the algorithm before tackling the implementation. Briefly, the algorithm consists of a precomputation phase in which element shape function derivatives $ \partial \mathbf{h} $ (and other quantities) and the system mass matrix $ \mathbf{M} $ are calculated, followed by a time-loop in which incremental solutions for the node displacements $ \mathbf{U} $ are found. During each step of the time-loop we:
\begin{enumerate}
\item Apply loads (displacements and/or forces) and boundary conditions to relevant nodal degrees of freedom
\item For each element compute
\begin{description}
\item[(a)] deformation gradient $ \mathbf{X} $ and right Cauchy-Green deformation tensor $ \mathbf{C} $
\item[(b)] linear strain-displacement matrix $ \mathbf{B}_L $
\item[(c)] second Piola-Kirchhoff stress $ \mathbf{S} $
\item[(d)] element nodal forces $ \mathbf{\tilde{F}} $, and add these forces to the total nodal forces $ \mathbf{F} $
\end{description}
\item For each node compute new displacements $ \mathbf{U} $ using the central difference method.
\end{enumerate}

@@

Memory storage is allowed through the use of textures, which essentially are structured collections of floating point values arranged as arrays. Shaders can read textyres from any location and as many time as they like. As such it is a concept very similar to regular arrays. However, the individual components of textures consist of up to four floating point values in the form of a vector. This comes from the fact that graphics processing units were initially designed to manipulate color images, so each component of a texture is in fact a vector holding the red, green, blue and alpha (transparency) values for the given pixel. Textures are therefore optimised for retrieving up to four floating point values in a single read.

When it comes to writing to GPU memory though, things are a bit more limited. Indeed, in contrast to texture reads, shaders can only write once into textures (at the end of the set of instructions). Morevoer, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location. Therefore, graphics API does not allow writing to random memory locations (this feature is called \emph{scattering}). Nethertheless, shaders can have multiple render targets. The write locations within each texture are still fixed though.

@@

	\subsubsection*{The choice of API}	
The first GPU implementation	of the TLED algorithm was carried out by \cite{Taylor07b,Taylor08}. They used the Cg language and had to reformulate the TLED algorithm presented to accomodate with the limited possibilities of this graphics API. Indeed, the major difference between this parallel implementation and a serial one is the writing of element nodal forces to memory at the end of a first kernel and their subsequent retrieval and summation during a second kernel. In a serial implementation nodal force contributions would most likely be added to a global sum as they are computed, rather than stored and summed in a second loop. But as stated earlier, this graphics API does not allow scattering, that is writing to random memory locations. Because of this limitation, element nodal forces cannot be directly added to the total nodal forces. Instead, the authors had to reformulate force summation as a gather operation in their implementation. They reported significant solution speed gains up to $ 16.8 \times$ over the CPU implementation. Thus, they were able to compute a simple cube model with up to $ 16\,000 $ tetrahedral elements in real-time.

@@

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks). \ON Yet, \OFF storing the force contributions in shared memory instead of global memory to read them later on via texture fetches \ON may \OFF outperform the need of computing several elements several times. \ON However, the efficiency of this single kernel approach depends on the nature of the mesh. Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements). This type of mesh is often encountered when dealing with the complex shapes of anatomical structures. Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective. 
\OFF

@@

\ON
This GPU implementation of the TLED algorithm was demonstrated publically at MICCAI 2007 during the talk given by Zeike Taylor \citep{Taylor07b}. A non-linear finite element analysis of a cube featuring brain-like properties was solved in real-time at $1\,430\,$Hz using $55\,566$ tetrahedral elements (see \fig{chap6:fig-demoMICCAI}).
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=11cm]{chapter6/demoMICCAI.pdf}
\end{center}
\caption[Real-time demo at MICCAI 2007]{Non-linear finite element algorithm solved in real-time at $1\,430\,$ Hz with $55\,566$ elements. A strain between 0 and $40\, \%$ is periodically applied to one face whilst the opposite face is fixed. The material properties were the following: cube's edge = $10\,$cm, Young modulus = $3\,000\,$ Pa, Poisson's ratio = 0.45, density = $1\,000\, kg.m^{-3}$ and the time step was $0.7\,$ms. The normal for each vertex was computed in real-time for a proper lightning and a skin texture was applied.} 
\label{chap6:fig-demoMICCAI}
\end{figure}
\OFF

@@

The tests were performed on a simple test scene featuring a cube under gravity. Hexahedral meshes with different resolutions from $1\,331$ to $29\,791$ nodes were used and the results are presented in figure \ref{chap6:fig-cube_and_curves}. In our standalone implementation, the second kernel not only accumulates nodal forces but also adds gravity and updates positions based on the central difference integration scheme. These operations are split into separated components in SOFA, in order to introduce more flexibility (such as applying additional forces or changing the integration algorithm). While this introduces no noticeable difference on a CPU-based simulation, when using the GPU it is more costly due to overheads in the CUDA API for the additional kernel launches. Although it reduces the performance by $8.4\, \%$ for large meshes, this could be optimised away by adding a kernel specific to a given combination of components.

	\subsection{Simulation of liver deformation} \label{chap6:liverSimu}
Finally, we performed simulations of manipulation of a liver model in order to demonstrate the feasibility of the procedure with more realistic organ geometries. The model geometry was based on the liver mesh available in the SOFA framework, which we smoothed and remeshed to improve uniformity of element sizes and node valencies. A mesh of 748 nodes and $2\,428$ tetrahedral elements was solvable in real-time. For anisotropic models a preferred material direction $\mathbf{a}_0 = [0\, 1\, 0] $ was used, corresponding to the vertical direction in \fig{chap6:fig-liver}. We note that in a real organ the preferred direction likely varies depending on the local vasculature and orientation of lobules \citep{Chui07}, but in the absence of experimental data for the present specimen the mentioned direction was assumed throughout. Moreover, on this point we note that different preferred directions may be specified for each element in the mesh, if such data are available.

@@

\TODO{Add movies online? If possible...}

@@

The first factor leads to a strong constraint on the quality of the mesh. Indeed, meshing a complex shape with good and uniform elements is not a straightforward task and rather often bad and very small elements are created. In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh. As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly \ON with great care \OFF to allow a decent simulation.

Material's properties are also a very influent factor for the determination of a critical time step. If the critical time step may be fairly large for very soft tissue (like brain), it is already divided by 2 when modelling liver. As we have seen, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. If it allows a time step twice as large, this relaxation introduces inaccuracies. If this is certainly acceptable for medical training simulators, the demand in precision for surgical planning or per-operating assistance would probably prevent such an approximation.

@@

	\subsection{Handling contacts}
The present work was entirely concerned with numerical solution of constitutive models in explicit FE analyses, however an equally important issue from the point of view of interactive simulation is the modelling of interaction between organs and virtual tools, and its possible impact on $ \Delta t_{cr} $ and the stability of the simulation. Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes. While this is beyond the scope of the present work, we note that a range of contact formulations are available, with two main types in common use: penalty methods and kinematic constraint methods. Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces. If excessively stiff springs are used, $ \Delta t_{cr} $ may indeed be reduced. \ON According to LS-Dyna documentation \citep{Hallquist06},  \OFF except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal. \ON However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system. \OFF Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $. \ON In theory, \OFF these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools. \ON It is worth nothing that \OFF the computational cost of either variety has not been investigated here.

\ON
Even if a stable contact modelling could be achieved, one element of crucial importance has been omitted so far: the collision detection process. Indeed, before even handling the interaction between two objects, their interaction needs to be detected. Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency. The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes. Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increase. One notable exception is the interaction between a soft deformable object and a fixed rigid surface where the collision detection is rapid.

@@

	\subsubsection{Conclusion}
As we showed in the last two chapters, the TLED algorithm is a very efficient fully non-linear finite element formulation. We also added viscoelasticity and anisotropy features with minimal impact on its performance. In particular, its GPU implementation allows non-linear analyses of structures in real-time using a high number of elements as never before.

However, the use of an explicit integration scheme imposes substantial constraints which limits TLED's range of applications. More specifically, the usage of the TLED algorithm in very interactive environments (as in the manipulation of a liver with a tool) is not recommended. The collision detection would either become the bottleneck of the simulation (if run at every time step) or fail to provide accurate detection of contacts (otherwise). In addition, the TLED algorithm is also not appropriate to simulate the deformation of rigid object as the critical time step would be too small to allow real-time computations. In constrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled. As an exemple, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.

Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
\OFF

>>>>

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation. If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation. The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation. This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA. 
\end{shortAbstract}

\section{Summary of the TLED formulation}
A complete description of the TLED algorithm for soft tissue simulation was given in the previous chapter. However, let us remind the main steps of the algorithm before tackling the implementation. Briefly, the algorithm consists of a pre-computation phase in which element shape function derivatives $ \partial \mathbf{h} $ (and other quantities) and the system mass matrix $ \mathbf{M} $ are calculated, followed by a time-loop in which incremental solutions for the node displacements $ \mathbf{U} $ are found. During each step of the time-loop we:
\begin{enumerate}
\item Apply loads (displacements and/or forces) and boundary conditions to relevant nodal degrees of freedom
\item For each element compute
\begin{description}
\item[(a)] deformation gradient $ \mathbf{X} $ and right Cauchy-Green deformation tensor $ \mathbf{C} $
\item[(b)] linear strain-displacement matrix $ \mathbf{B}_L $
\item[(c)] second Piola-Kirchhoff stress $ \mathbf{S} $
\item[(d)] element nodal forces $ \mathbf{\tilde{F}} $, and add these forces to the total nodal forces $ \mathbf{F} $
\end{description}
\item For each node compute new displacements $ \mathbf{U} $ using the central difference method.
\end{enumerate}

@@

Memory storage is allowed through the use of textures, which essentially are structured collections of floating point values arranged as arrays. Shaders can read textures from any location and as many time as they like. As such it is a concept very similar to regular arrays. However, the individual components of textures consist of up to four floating point values in the form of a vector. This comes from the fact that graphics processing units were initially designed to manipulate color images, so each component of a texture is in fact a vector holding the red, green, blue and alpha (transparency) values for the given pixel. Textures are therefore optimised for retrieving up to four floating point values in a single read.

When it comes to writing to GPU memory though, things are a bit more limited. Indeed, in contrast to texture reads, shaders can only write once into textures (at the end of the set of instructions). Moreover, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location. Therefore, graphics API does not allow writing to random memory locations (this feature is called \emph{scattering}). Nevertheless, shaders can have multiple render targets. The write locations within each texture are still fixed though.

@@

	\subsubsection*{The choice of API}	
The first GPU implementation	of the TLED algorithm was carried out by \cite{Taylor07b,Taylor08}. They used the Cg language and had to reformulate the TLED algorithm presented to accommodate with the limited possibilities of this graphics API. Indeed, the major difference between this parallel implementation and a serial one is the writing of element nodal forces to memory at the end of a first kernel and their subsequent retrieval and summation during a second kernel. In a serial implementation nodal force contributions would most likely be added to a global sum as they are computed, rather than stored and summed in a second loop. But as stated earlier, this graphics API does not allow scattering, that is writing to random memory locations. Because of this limitation, element nodal forces cannot be directly added to the total nodal forces. Instead, the authors had to reformulate force summation as a gather operation in their implementation. They reported significant solution speed gains up to $ 16.8 \times$ over the CPU implementation. Thus, they were able to compute a simple cube model with up to $ 16\,000 $ tetrahedral elements in real-time.

@@

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks).  Yet,  storing the force contributions in shared memory instead of global memory to read them later on via texture fetches  may  outperform the need of computing several elements several times.  However, the efficiency of this single kernel approach depends on the nature of the mesh. Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements). This type of mesh is often encountered when dealing with the complex shapes of anatomical structures. Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective.

@@

This GPU implementation of the TLED algorithm was demonstrated publically at MICCAI 2007 during the talk given by Zeike Taylor \citep{Taylor07b}. A non-linear finite element analysis of a cube featuring brain-like properties was solved in real-time at $1\,430\,$Hz using $55\,566$ tetrahedral elements (see \fig{chap6:fig-demoMICCAI}).
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=11cm]{chapter6/demoMICCAI.pdf}
\end{center}
\caption[Real-time demo at MICCAI 2007]{Non-linear finite element algorithm solved in real-time at $1\,430\,$ Hz with $55\,566$ elements. A strain between 0 and $40\, \%$ is periodically applied to one face whilst the opposite face is fixed. The material properties were the following: cube's edge = $10\,$cm, Young modulus = $3\,000\,$ Pa, Poisson's ratio = 0.45, density = $1\,000\, kg.m^{-3}$ and the time step was $0.7\,$ms. The normal for each vertex was computed in real-time for a proper lightning and a skin texture was applied.} 
\label{chap6:fig-demoMICCAI}
\end{figure}

@@

The tests were performed on a simple test scene featuring a cube under gravity. Hexahedral meshes with different resolutions from $1\,331$ to $29\,791$ nodes were used and the results are presented in \fig{chap6:fig-cube_and_curves}. In our standalone implementation, the second kernel not only accumulates nodal forces but also adds gravity and updates positions based on the central difference integration scheme. These operations are split into separated components in SOFA, in order to introduce more flexibility (such as applying additional forces or changing the integration algorithm). While this introduces no noticeable difference on a CPU-based simulation, when using the GPU it is more costly due to overheads in the CUDA API for the additional kernel launches. Although it reduces the performance by $8.4\, \%$ for large meshes, this could be optimised away by adding a kernel specific to a given combination of components.

	\subsection{Simulation of liver deformation} \label{chap6:liverSimu}
Finally, we performed simulations of manipulation of a liver model in order to demonstrate the feasibility of the procedure with more realistic organ geometries. The model geometry was based on the liver mesh available in the SOFA framework, which we smoothed and re-meshed to improve uniformity of element sizes and node valencies. A mesh of 748 nodes and $2\,428$ tetrahedral elements was solvable in real-time. For anisotropic models a preferred material direction $\mathbf{a}_0 = [0\, 1\, 0] $ was used, corresponding to the vertical direction in \fig{chap6:fig-liver}. We note that in a real organ the preferred direction likely varies depending on the local vasculature and orientation of lobules \citep{Chui07}, but in the absence of experimental data for the present specimen the mentioned direction was assumed throughout. Moreover, on this point we note that different preferred directions may be specified for each element in the mesh, if such data are available.

@@

@@

The first factor leads to a strong constraint on the quality of the mesh. Indeed, meshing a complex shape with good and uniform elements is not a straightforward task and rather often bad and very small elements are created. In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh. As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and re-meshed properly  with great care  to allow a decent simulation.

Material's properties also have an influence on the determination of a critical time step. If the critical time step may be fairly large for very soft tissue (like brain), it is already divided by 2 when modelling liver. As we have seen, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. If it allows a time step twice as large, this relaxation introduces inaccuracies. If this is certainly acceptable for medical training simulators, the demand in precision for surgical planning or per-operating assistance would probably prevent such an approximation.

@@

	\subsection{Handling contacts}
The present work was entirely concerned with numerical solution of constitutive models in explicit FE analyses, however an equally important issue from the point of view of interactive simulation is the modelling of interaction between organs and virtual tools, and its possible impact on $ \Delta t_{cr} $ and the stability of the simulation. Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes. While this is beyond the scope of the present work, we note that a range of contact formulations are available, with two main types in common use: penalty methods and kinematic constraint methods. Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces. If excessively stiff springs are used, $ \Delta t_{cr} $ may indeed be reduced.  According to LS-Dyna documentation \citep{Hallquist06},   except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal.  However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system.  Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.  In theory,  these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.  It is worth nothing that  the computational cost of either variety has not been investigated here.

Even if a stable contact modelling could be achieved, one element of crucial importance has been omitted so far: the collision detection process. Indeed, before even handling the interaction between two objects, their interaction needs to be detected. Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency. The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes. Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increase. One notable exception is the interaction between a soft deformable object and a fixed rigid surface where the collision detection is rapid.

@@

\section{Conclusion}

As we showed in the last two chapters, the TLED algorithm is a very efficient fully non-linear finite element formulation. We also added viscoelasticity and anisotropy features with minimal impact on its performance. In particular, its GPU implementation allows non-linear analyses of structures in real-time using a high number of elements as never before.

However, the use of an explicit integration scheme imposes substantial constraints which limits TLED's range of applications. More specifically, the usage of the TLED algorithm in very interactive environments (as in the manipulation of a liver with a tool) is not recommended. The collision detection would either become the bottleneck of the simulation (if run at every time step) or fail to provide accurate detection of contacts (otherwise). In addition, the TLED algorithm is also not appropriate to simulate the deformation of rigid object as the critical time step would be too small to allow real-time computations. In contrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled. As an example, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.

Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, my efficient re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. My contribution was once again to implement these enhancements on GPU. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. Over the course of my PhD, this work has led to three main publications \citep{Comas2008,Taylor2008,Taylor2009}.
